Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-06 revised_manuscript.Rmd
update

We tested 29 deaf adult signers, 15 with exposure to TİD from birth (i.e., native signers), 14 with exposure between ages 3 and 17 (i.e., late signers) (*M*\textsubscript{Age of acquisition} = 7.7 years, *SD* = 3.3 years). The criterion to be classified as a late signer was to have hearing and non-signing caregivers. All the signers satisfying this criterion were regarded as late signers \footnote{Note that we are essentially comparing deaf of deaf and deaf of hearing adult signers. Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among late signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}}. All participants indicated that their preferred language of communication was TİD which they used for most contexts. Detailed information about the linguistic and educational background of the participants can be found in \ref{section:a}.

All participating native signers (7 males, 8 females) were born into deaf families and started acquiring TİD from at least one of their caregivers after birth. The parents of 14 participating native signers were deaf. One native signer had a deaf father and a hearing mother. In addition to TİD, the home language for some native signers (n = 4) in early childhood included the use of home sign\footnote{Home signs are gesture-based communication systems that emerge between the deaf children and their primary caregivers. Although home sign communication may have internal consistency, they are not fully-flegged linguistic systems like natural languages}. The age of the native participants ranged between 18 and 35 (*M* = 26.4 years, *SD* = 5.0 years). This also corresponded to the total average years of TİD use for native signers. All native signers were pre-lingually deaf, that is they were either born deaf or they had become deaf before the age of 3, with the exception of one signer who became deaf between the ages of 4 and 7. Native signers reported to complete at least high school education and their level of education (LOE) was between 12 and 16 years (*M* = 12.7 years, *SD* = 1.5 years). In addition, all native participants attended at least two deaf schools, one being primary and the other middle school. 10 native participants also attended a deaf school for their high school education. Based on a self-reported 5-point scale of proficiency (1: poor, 5: proficient), native participants declared that they have high receptive and productive skills in TİD (*M* = 5 points, *SD* = 0 points). As for their proficiency in (spoken) Turkish, their self-reported rankings were on average 3.7 (*SD* = 0.7), 3.7 (*SD* = 0.7), and 2.6 (*SD* = 1.2) for reading, writing, and speaking abilities, respectively.

@@

[^3]: As in other sign languages, in TİD, the lexical meaning of signs has an effect on the morphological grouping. For instance, usually (di)transitive agreement verbs denote some sort of physical or metaphorical transfer of an object or abstract notion [@meir1999]. As for plain verbs, both phonology of a sign and its semantics have an effect on morphology. For instance, \textsc{eat} in TİD is signed nearby the mouth and since it needs to remain in that phonological location, it is not suitable for agreement and thus it is regarded as a plain verb. Likewise the experiencer verb \textsc{love} is signed on the torso and as such it does not take part in agreement. These generalizations noted we should also be cautious in labelling a verb as categorially one type or the other. For instance, some intransitive plain verbs like \textsc{fall} or \textsc{jump} in TİD may display spatial person agreement with the single argument in the clause [@gokgoz2020]. See @lourenço2018 also for a similar discussion in Brazilian Sign Language.}.

>>>>

We tested 29 deaf adult signers, 15 with exposure to TİD from birth (i.e., native signers), 14 with exposure between ages 3 and 17 (i.e., late signers) (*M*\textsubscript{Age of acquisition} = 7.7 years, *SD* = 3.3 years). The criterion to be classified as a late signer was to have hearing and non-signing caregivers. All the signers satisfying this criterion were regarded as late signers \footnote{Note that we are essentially comparing deaf of deaf and deaf of hearing adult signers. Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among late signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}.}. All participants indicated that their preferred language of communication was TİD which they used for most contexts. Detailed information about the linguistic and educational background of the participants can be found in \ref{section:a}.

All participating native signers (7 males, 8 females) were born into deaf families and started acquiring TİD from at least one of their caregivers after birth. The parents of 14 participating native signers were deaf. One native signer had a deaf father and a hearing mother. In addition to TİD, the home language for some native signers (n = 4) in early childhood included the use of home sign\footnote{Home signs are gesture-based communication systems that emerge between the deaf children and their primary caregivers. Although home sign communication may have internal consistency, they are not fully-flegged linguistic systems like natural languages.}. The age of the native participants ranged between 18 and 35 (*M* = 26.4 years, *SD* = 5.0 years). This also corresponded to the total average years of TİD use for native signers. All native signers were pre-lingually deaf, that is they were either born deaf or they had become deaf before the age of 3, with the exception of one signer who became deaf between the ages of 4 and 7. Native signers reported to complete at least high school education and their level of education (LOE) was between 12 and 16 years (*M* = 12.7 years, *SD* = 1.5 years). In addition, all native participants attended at least two deaf schools, one being primary and the other middle school. 10 native participants also attended a deaf school for their high school education. Based on a self-reported 5-point scale of proficiency (1: poor, 5: proficient), native participants declared that they have high receptive and productive skills in TİD (*M* = 5 points, *SD* = 0 points). As for their proficiency in (spoken) Turkish, their self-reported rankings were on average 3.7 (*SD* = 0.7), 3.7 (*SD* = 0.7), and 2.6 (*SD* = 1.2) for reading, writing, and speaking abilities, respectively.

@@

[^3]: As in other sign languages, in TİD, the lexical meaning of signs has an effect on the morphological grouping. For instance, usually (di)transitive agreement verbs denote some sort of physical or metaphorical transfer of an object or abstract notion [@meir1999]. As for plain verbs, both phonology of a sign and its semantics have an effect on morphology. For instance, \textsc{eat} in TİD is signed nearby the mouth and since it needs to remain in that phonological location, it is not suitable for agreement and thus it is regarded as a plain verb. Likewise the experiencer verb \textsc{love} is signed on the torso and as such it does not take part in agreement. These generalizations noted we should also be cautious in labelling a verb as categorially one type or the other. For instance, some intransitive plain verbs like \textsc{fall} or \textsc{jump} in TİD may display spatial person agreement with the single argument in the clause [@gokgoz2020]. See @lourenço2018 also for a similar discussion in Brazilian Sign Language.

### 2022-08-06 revised_manuscript.tex
update

We tested 29 deaf adult signers, 15 with exposure to TİD from birth
(i.e., native signers), 14 with exposure between ages 3 and 17 (i.e.,
late signers) (\emph{M}\textsubscript{Age of acquisition} = 7.7 years,
\emph{SD} = 3.3 years). The criterion to be classified as a late signer
was to have hearing and non-signing caregivers. All the signers
satisfying this criterion were regarded as late signers
\footnote{Note that we are essentially comparing deaf of deaf and deaf of hearing adult signers. Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among late signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}}.
All participants indicated that their preferred language of
communication was TİD which they used for most contexts. Detailed
information about the linguistic and educational background of the
participants can be found in \ref{section:a}.

All participating native signers (7 males, 8 females) were born into
deaf families and started acquiring TİD from at least one of their
caregivers after birth. The parents of 14 participating native signers
were deaf. One native signer had a deaf father and a hearing mother. In
addition to TİD, the home language for some native signers (n = 4) in
early childhood included the use of home
sign\footnote{Home signs are gesture-based communication systems that emerge between the deaf children and their primary caregivers. Although home sign communication may have internal consistency, they are not fully-flegged linguistic systems like natural languages}.
The age of the native participants ranged between 18 and 35 (\emph{M} =
26.4 years, \emph{SD} = 5.0 years). This also corresponded to the total
average years of TİD use for native signers. All native signers were
pre-lingually deaf, that is they were either born deaf or they had
become deaf before the age of 3, with the exception of one signer who
became deaf between the ages of 4 and 7. Native signers reported to
complete at least high school education and their level of education
(LOE) was between 12 and 16 years (\emph{M} = 12.7 years, \emph{SD} =
1.5 years). In addition, all native participants attended at least two
deaf schools, one being primary and the other middle school. 10 native
participants also attended a deaf school for their high school
education. Based on a self-reported 5-point scale of proficiency (1:
poor, 5: proficient), native participants declared that they have high
receptive and productive skills in TİD (\emph{M} = 5 points, \emph{SD} =
0 points). As for their proficiency in (spoken) Turkish, their
self-reported rankings were on average 3.7 (\emph{SD} = 0.7), 3.7
(\emph{SD} = 0.7), and 2.6 (\emph{SD} = 1.2) for reading, writing, and
speaking abilities, respectively.

@@

As for the distribution of the null markers, approximately 90\% of the
plain verbs were used to maintain a previous referent, making these
constructions least implicit REs used for TİD\footnote{As in other sign
  languages, in TİD, the lexical meaning of signs has an effect on the
  morphological grouping. For instance, usually (di)transitive agreement
  verbs denote some sort of physical or metaphorical transfer of an
  object or abstract notion (Meir, 1999). As for plain verbs, both
  phonology of a sign and its semantics have an effect on morphology.
  For instance, \textsc{eat} in TİD is signed nearby the mouth and since
  it needs to remain in that phonological location, it is not suitable
  for agreement and thus it is regarded as a plain verb. Likewise the
  experiencer verb \textsc{love} is signed on the torso and as such it
  does not take part in agreement. These generalizations noted we should
  also be cautious in labelling a verb as categorially one type or the
  other. For instance, some intransitive plain verbs like \textsc{fall}
  or \textsc{jump} in TİD may display spatial person agreement with the
  single argument in the clause (Gökgöz and Sevgi, 2020). See Lourenço
  and Wilbur (2018) also for a similar discussion in Brazilian Sign
  Language.\}.}. Agreement verbs and CA were the following implicit
markers of reference, 0.77 and 0.72 of which were used for maintained
contexts, respectively. This contrasts with some of the research
(Czubek, 2017) in which agreement verbs ranked lowest on scale,
indicating that these constructions were more implicit compared to plain
verbs. However, there seems to be a common ground for the position of CA
markers which have been claimed to rank higher than other null markers
as in the present study (Barberà and Massó, 2009; Czubek, 2017;
Frederiksen and Mayberry, 2016). In Figure \ref{fig:scale}, we report
the distribution of REs in TİD as a function of referent accessibility
and compare our accessibility hierarchy with that of (Frederiksen and
Mayberry, 2016) for ASL. In the hierarchies, the leftmost REs (e.g.,
modified and bare nouns for TİD) are used when the accessibility is low.
If the referent is already active in the addressee's mind and its
accessibility is high, then the rightmost markers (e.g., agreement and
plain verbs for TİD) are selected.

>>>>

We tested 29 deaf adult signers, 15 with exposure to TİD from birth
(i.e., native signers), 14 with exposure between ages 3 and 17 (i.e.,
late signers) (\emph{M}\textsubscript{Age of acquisition} = 7.7 years,
\emph{SD} = 3.3 years). The criterion to be classified as a late signer
was to have hearing and non-signing caregivers. All the signers
satisfying this criterion were regarded as late signers
\footnote{Note that we are essentially comparing deaf of deaf and deaf of hearing adult signers. Although we did not set a cut-off point for nativeness, all the deaf of hearing participants indicated that they learned TİD after age 3. In order to capture in-group variation among late signers, we run an additional supplementary analysis in which we observed very wide credible intervals due to the small group size, rendering the results uninformative. For the analysis, see \ref{section:d}.}.
All participants indicated that their preferred language of
communication was TİD which they used for most contexts. Detailed
information about the linguistic and educational background of the
participants can be found in \ref{section:a}.

All participating native signers (7 males, 8 females) were born into
deaf families and started acquiring TİD from at least one of their
caregivers after birth. The parents of 14 participating native signers
were deaf. One native signer had a deaf father and a hearing mother. In
addition to TİD, the home language for some native signers (n = 4) in
early childhood included the use of home
sign\footnote{Home signs are gesture-based communication systems that emerge between the deaf children and their primary caregivers. Although home sign communication may have internal consistency, they are not fully-flegged linguistic systems like natural languages.}.
The age of the native participants ranged between 18 and 35 (\emph{M} =
26.4 years, \emph{SD} = 5.0 years). This also corresponded to the total
average years of TİD use for native signers. All native signers were
pre-lingually deaf, that is they were either born deaf or they had
become deaf before the age of 3, with the exception of one signer who
became deaf between the ages of 4 and 7. Native signers reported to
complete at least high school education and their level of education
(LOE) was between 12 and 16 years (\emph{M} = 12.7 years, \emph{SD} =
1.5 years). In addition, all native participants attended at least two
deaf schools, one being primary and the other middle school. 10 native
participants also attended a deaf school for their high school
education. Based on a self-reported 5-point scale of proficiency (1:
poor, 5: proficient), native participants declared that they have high
receptive and productive skills in TİD (\emph{M} = 5 points, \emph{SD} =
0 points). As for their proficiency in (spoken) Turkish, their
self-reported rankings were on average 3.7 (\emph{SD} = 0.7), 3.7
(\emph{SD} = 0.7), and 2.6 (\emph{SD} = 1.2) for reading, writing, and
speaking abilities, respectively.

@@

As for the distribution of the null markers, approximately 90\% of the
plain verbs were used to maintain a previous referent, making these
constructions least implicit REs used for TİD\footnote{As in other sign
  languages, in TİD, the lexical meaning of signs has an effect on the
  morphological grouping. For instance, usually (di)transitive agreement
  verbs denote some sort of physical or metaphorical transfer of an
  object or abstract notion (Meir, 1999). As for plain verbs, both
  phonology of a sign and its semantics have an effect on morphology.
  For instance, \textsc{eat} in TİD is signed nearby the mouth and since
  it needs to remain in that phonological location, it is not suitable
  for agreement and thus it is regarded as a plain verb. Likewise the
  experiencer verb \textsc{love} is signed on the torso and as such it
  does not take part in agreement. These generalizations noted we should
  also be cautious in labelling a verb as categorially one type or the
  other. For instance, some intransitive plain verbs like \textsc{fall}
  or \textsc{jump} in TİD may display spatial person agreement with the
  single argument in the clause (Gökgöz and Sevgi, 2020). See Lourenço
  and Wilbur (2018) also for a similar discussion in Brazilian Sign
  Language.}. Agreement verbs and CA were the following implicit markers
of reference, 0.77 and 0.72 of which were used for maintained contexts,
respectively. This contrasts with some of the research (Czubek, 2017) in
which agreement verbs ranked lowest on scale, indicating that these
constructions were more implicit compared to plain verbs. However, there
seems to be a common ground for the position of CA markers which have
been claimed to rank higher than other null markers as in the present
study (Barberà and Massó, 2009; Czubek, 2017; Frederiksen and Mayberry,
2016). In Figure \ref{fig:scale}, we report the distribution of REs in
TİD as a function of referent accessibility and compare our
accessibility hierarchy with that of (Frederiksen and Mayberry, 2016)
for ASL. In the hierarchies, the leftmost REs (e.g., modified and bare
nouns for TİD) are used when the accessibility is low. If the referent
is already active in the addressee's mind and its accessibility is high,
then the rightmost markers (e.g., agreement and plain verbs for TİD) are
selected.
